## CSAMA

In this demo, we'll be analyzing 5K-plex Xenium (10x Genomics) 
data on a tissue section from a human breast cancer biopsy, 
derived from [here](https://www.10xgenomics.com/datasets/xenium-prime-ffpe-human-breast-cancer).
These include the following elements:

- two `image`s (histology and immunofluorescence)
- two `shape`s (nuclei and membrane boundaries)
- one `table` (gene $\times$ cell data)
- one `point` (RNA transcripts)

### load data and libraries

A size-reduced example dataset has been deposited on Zenodo. 
We'll first retrieve the compressed `SpatialData` `.zarr` store together
with a reference scRNA-seq dataset from [Janesick et al.](https://www.nature.com/articles/s41467-023-43458-x), 
decompress both archives, and stash their on-disk file paths for later:

```{r load-data}
# retrieve SpatialData .zarr
url_zarr <- file.path(
    "https://zenodo.org/records/20180722/files",
    "BC_10x_xenium_sdata.zarr.zip?download=1")

zip_zarr <- tempfile("sd", fileext=".zarr.zip")
download.file(url_zarr, zip_zarr, quiet=TRUE)

# decompress .zarr
nm_zarr <- unzip(zip_zarr, exdir=dirname(zip_zarr))
zs <- grep("zarr$", dirname(nm_zarr), value=TRUE)

# retrieve scRNA-seq reference
url_scrna <- file.path(
    "https://zenodo.org/records/20178278/files",
    "BC_scRNAseq_subset.h5ad.zip?download=1")

zip_scrna <- tempfile("scrna", fileext=".h5ad.zip")
download.file(url_scrna, zip_scrna, quiet=TRUE)

# decompress .h5ad
nm_scrna <- unzip(zip_scrna, exdir=dirname(zip_scrna))
h5 <- grep("h5ad$", nm_scrna, value=TRUE)

# show retrieved files
cat(
    basename(zs), "\n",
    basename(h5), "\n"
)
```

```{r show-zarr, echo=FALSE}
htmltools::pre(
    style="max-height:333px; overflow-y:auto;",
    paste(
        capture.output(fs::dir_tree(zs, recurse=2))[-1],
        collapse="\n"))
```

```{r load-libs, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(anndataR)
library(scrapper)
library(patchwork)
library(reticulate)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
library(SingleCellExperiment)
library(ggnewscale)
```

### image visualization

Let's load up the `SpatialData` object in `R`. 
```{r}
sd <- readSpatialData(zs)
sd
```

#### histopathology

Let's start out by visualization the hematoxylin and eosin (H&E) staining.
H\&E provides a high-contrast visualization of tissue microarchitecture and 
cellular morphology through differential staining of nuclear and cytoplasmic 
components. To date, whole-slide H\&E images (WSIs) remain the clinical gold 
standard for cancer diagnosis, including tumor grading, staging, and prognosis.

```{r plt-he}
plotSpatialData() + plotImage(sd)
```

We can also zoom into a particular region by specifying a bounding box and
using `crop()` to subset the data accordingly:

```{r plt-he-box}
bb <- list(
    xmin=15e3, xmax=18e3, 
    ymin=44e3, ymax=46e3)
sp <- crop(sd, bb)
plotSpatialData() + plotImage(sp) 
```

#### immunofluorescence

In addition to the H\&E, the dataset includes 4 immunofluorescence (IF) markers:

* DAPI (nuclei) marks all cell nuclei and is used for cell segmentation
* ATP1A1/CD45/E-Cadherin (membrane) marks cell boundaries and major lineages
* 18S (cytoplasm) marks the cell body/cytoplasmic RNA and overall cell extent
* αSMA/Vimentin (stroma) marks mesenchymal/stromal tissue, including fibroblasts

These channels are part of the 10x segmentation kit and are used in the [multimodal cell segmentation workflow](https://cdn.10xgenomics.com/image/upload/v1754601291/support-documents/CG000750_XeniumInSitu_CellSegmentation_TechNote_RevB.pdf).

```{r if-chs}
img <- image(sd, 2)
as.vector(channels(img))
```

The underlying image is a 3D array where the first dimension are channels, 
and the second and third dimensions are height and width; we can verify this
by viewing the `axes()` specified in the *.zattrs* associated with the image:

```{r if-dim}
do.call(rbind, axes(img))
cat(dim(img))
```

Let's visualize the four-plex composite image overlaying these markers;
for consistent visualization, we also specify colors to use for each channel:

```{r plt-if-all}
pal <- c("blue", "green", "cyan", "magenta")
chs <- names(pal) <- channels(img)
plotSpatialData() + plotImage(sd, 2, ch=chs, c=pal[chs])
```

Alternatively, we can visualize each channel separately; 
here, using the previously defined smaller region:

```{r plt-if-one}
ps <- lapply(chs, \(.) plotSpatialData() + plotImage(sp, 2, ch=., c=pal[.]) + ggtitle(.))
wrap_plots(ps, nrow=2) & theme_void() & theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

#### segmentation masks

We can also zoom in further to inspect how well cell segmentation boundaries 
align with histopathology and the immunofluorescence signals used to get them:

```{r plt-if-seg, fig.width=11, fig.height=6}
bb <- list(
    xmin=15e3, xmax=17e3,
    ymin=42e3, ymax=45e3)
sq <- crop(sd, bb)
plotSpatialData() + 
    plotImage(sq, 1) +
    plotShape(sq, fill=NA, col="black", linewidth=1/5) +
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, fill=NA, col="white", linewidth=1/5)
```

We can also go a step further, visualizing both cell nuclei (in gold) 
and membrane (in grey) boundaries togeher on an ever smaller region:

```{r crop-1}
bb <- list(
    xmin=15e3, xmax=16e3,
    ymin=43e3, ymax=44e3)
sq <- crop(sd, bb)
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, 1, fill=NA, col="grey", linewidth=1/3) +
    plotShape(sq, 2, fill=NA, col="gold", linewidth=1/3)
```

#### transcripts 

Going all the way, let's zoom in to a couple cells in order to visualize 
4,000 individual transcript molecules as well; we'll color these by whether
or not they have been determined to `overlaps_nucleus` (according to 10x):

```{r}
bb <- list(
    xmin=15200, xmax=15800,
    ymin=43200, ymax=43800)
sq <- crop(sd, bb)
point(sq) <- mutate(point(sq), overlaps_nucleus=as.logical(overlaps_nucleus))
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, 1, fill=NA, col="grey", linewidth=1/3) +
    plotShape(sq, 2, fill=NA, col="gold", linewidth=1/3) +
    plotPoint(sq, n=4e3, col="overlaps_nucleus", size=0) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    scale_color_manual(values=c("grey", "red"))
```

Going back to a larger scale...

```{r fig.width=8, fig.height=4}
#| code-fold: true
bb <- list(xmin=16e3, xmax=18e3, ymin=45e3, ymax=46e3)
sq <- crop(sd[c("images", "points"), list(2, 1)], bb)
plotSpatialData() + 
    plotImage(sq, ch="DAPI", c="purple") +
    plotPoint(sq, n=8e3, col="ivory", size=0) +
    theme_void() + theme(legend.position="none")
```

### segmentation qc
Visual inspection is essential; however, accurately assessing transcript
assignment would in principle require inspection of transcript localization in
3D across thousands of genes, which quickly becomes infeasible for modern
high-plex panels. The [`SegTraQ`](https://github.com/LazDaria/SegTraQ) package provides a colloection of metrics for **Seg**mentation and **Tr**anscript **A**ssignment **Q**uality Control. 
While `SegTraQ` is implemented in Python, `reticulate` enables seamless
interoperability between Python and R. We can therefore compute QC metrics in
Python and subsequently switch back to R for visualization and downstream
exploration.

#### running segtraq in python

We can instruct our R session to use an existing Python environment with our
required dependencies (here, `SegTraQ`) via `reticulate`.
Below we demonstrate two common approaches:

- a **conda/mamba** environment
- a lightweight **uv** virtual environment

::: {.panel-tabset}

##### conda / mamba

This environment was created previously from the command line:

```bash
conda create -n segtraq python=3.11
conda activate segtraq
pip install segtraq
```

We can then activate it from R using `reticulate`:

```{r py-env-conda, eval=FALSE}
library(reticulate)

# path to conda binary
bin <- "~/software/mambaforge/bin/conda" 
options(reticulate.conda_binary=bin)

# activate environment
use_condaenv("segtraq")
```

##### uv

Alternatively, we can use a local virtual environment created with `uv` (install via `brew install uv`).

This environment was created previously from the command line:

```bash
uv venv .venv --python 3.11
source .venv/bin/activate
uv pip install segtraq
```

We can then point `reticulate` directly to the Python executable:

```{r py-env-uv}
library(reticulate)

venv_path <- file.path(
    getwd(),
    ".venv/bin/python"
)

use_python(venv_path, required = TRUE)
```

:::

##### load data and libraries
Once configured, required Python modules can be imported:

```{python py-opt, echo=FALSE}
import joblib
import logging
import warnings
import spatialdata as sd
import segtraq
import anndata as ad
import spatialdata_plot
import matplotlib.pyplot as plt

warnings.filterwarnings("ignore")
joblib.parallel.DEFAULT_BACKEND = "threading"

logging.getLogger("spatialdata").propagate = False
logging.getLogger("spatialdata").setLevel(logging.ERROR)

logging.getLogger("spatialdata_plot").propagate = False
logging.getLogger("spatialdata_plot").setLevel(logging.ERROR)
```

We first load the `SpatialData` object from the `.zarr` store in Python:
```{python segtraq-load}
sdata = sd.read_zarr(r.zs)
sdata
```

##### compute SegTraQ metrics

`SegTraQ` organizes metrics into six modules; here, we begin with the
baseline (`bl`) module, which computes per-cell and per-sample summary statistics
such as transcript counts, detected genes, cell area, and fractions of
unassigned transcripts.

Metrics can either be computed individually, e.g. via `st.bl.transcripts_per_cell()`
or jointly by running the full `bl` module with `st.run_baseline()`.

```{python segtraq-baseline}
st = segtraq.SegTraQ(sdata,
    tables_centroid_x_key="centroid_x",
    tables_centroid_y_key="centroid_y",
    images_key="morphology_focus")

st.run_baseline()
```

Of the six `SegTraQ` modules, only the supervised (`sp`) module requires scRNA-seq
reference data. Label transfer can either be performed using a lightweight
`SegTraQ` helper, or supplied by the user directly.

Here, we use the scRNA-seq reference from Janesick et al. downloaded above
and transfer labels to the spatial data using a `SegTraQ` helper.

```{python segtraq-labeltransfer}
ref = ad.read_h5ad(r.h5)

st.run_label_transfer(
    ref, 
    ref_cell_type="celltype", 
    ref_ensemble_key=None, 
    query_ensemble_key=None)
```

We use this reference to derive positive and negative cell type markers,
i.e. genes expected to be present or absent in a given cell type, and pass
these to the `sp` module to identify contaminated cells. For demonstration,
we compute two metrics from the `sp` module.

```{python segtraq-supervised}
#subset to common genes
adata = st.sdata.tables["table"]
common_genes = adata.var_names[adata.var_names.isin(ref.var_names)]
ref = ref[:, common_genes].copy()

markers = segtraq.markers_from_reference(
    ref,
    cell_type_key="celltype",
    n_jobs=16,
)

_ = st.sp.neighbor_contamination(
    markers=markers,
    cell_type_key="transferred_cell_type"
)

_ = st.sp.marker_purity(
    markers=markers,
    cell_type_key="transferred_cell_type"
)

```

In the absence of scRNA-seq data, `SegTraQ` provides the region similarity (`rs`) module, which identifies contaminated cells based on the expression similarity between subcellular regions and the local neighborhood. For demonstration, we one metric from the `rs` module.

```{python segtraq-region-similarity}
st.rs.similarity_border_neighborhood().head()
```

We can visualize the H&E image, the immunofluorescence channel DAPI, and the cell boundaries colored by the transferred cell type label or `SegTraQ` metrics. 

```{python segtraq-plot}
#| code-fold: true

# define color map
col_celltype = {
    "T": "#fb8072",
    "B": "#bc80bd",
    "macro": "#910290",
    "dendritic": "#fdb462",
    "mast": "#959059",
    "perivas": "#fed9a6",
    "endo": "#a6cee3",
    "myoepi": "#2782bb",
    "DCIS1": "#3c7761",
    "DCIS2": "#66a61e",
    "tumor": "#66c2a5",
    "stromal": "#d45943",
    "nan": "#808080",
}

# compute per-cell color
labels = sdata.tables["table"].obs["transferred_cell_type"].unique().astype(str).tolist()
cols = [col_celltype[lab] for lab in labels]

fg, ax = plt.subplots(2, 3, figsize=(14, 8), constrained_layout=True)
fg.subplots_adjust(wspace=0.5)

# plot histopathology
st.sdata.pl.render_images(
    "he_image"
).pl.show(
    ax=ax[0, 0], 
    title="H&E")

# plot immunofluorescence
st.sdata.pl.render_images(
    "morphology_focus", 
    channel="DAPI"
).pl.show(
    ax=ax[0, 1], 
    title="DAPI")

# plot cell boundaries colored by cell type
st.sdata.pl.render_shapes(
    element="cell_boundaries",
    color="transferred_cell_type",
    palette=cols,
    groups=labels,
    method="matplotlib"
).pl.show(
    ax=ax[0, 2], 
    colorbar=True,
    title="Cell type")

# plot cell boundaries colored by elongation
st.sdata.pl.render_shapes(
    element="cell_boundaries",
    color="elongation",
    method="matplotlib"
).pl.show(
    ax=ax[1, 0], 
    colorbar=True,
    title="Elongation")

# plot cell boundaries colored by transcript count
st.sdata.pl.render_shapes(
    element="cell_boundaries",
    color="transcript_count",
    method="matplotlib"
).pl.show(
    ax=ax[1, 1], 
    colorbar=True,
    title="Transcript count")
    
# plot cell boundaries colored by negative marker counts
st.sdata.pl.render_shapes(
    element="cell_boundaries",
    color="negative_marker_contamination_counts",
    method="matplotlib"
).pl.show(
    ax=ax[1, 2], 
    colorbar=True,
    title="Negative marker (contamination) count")
```

```{r echo=FALSE}
knitr::opts_chunk$set(eval=TRUE)
```

After computing required QC metrics in Python, we can move on to explore and visualize these in more detail in `R`.

#### exploration of QC metrics in R
Let's retrieve the 
`table`, including `SegTraQ` results, from Python, and transfer it to the 
`SpatialData` object in R. We set the `table` to annotate cell
segmentation boundaries using `setTable()`; `ik` specifies the `spatialdata`'s
`instance_key` used to link observations between `table` and spatial element.

```{r transfer-python-table}
# set Python table to annotate 
# cell segmentation boundaries
se <- py$adata$as_SingleCellExperiment()

# remove old table
sd <- sd[c("images", "points", "shapes")]
(sd <- setTable(sd, y=se, 
    i="cell_boundaries", 
    ik="cell_id"))
```

Let's look at a summary of baseline metrics:
```{r inspect-baseline-metrics}
tbl <- table(sd)

num_cells <- metadata(tbl)$num_cells

p_unassigned <- metadata(tbl)$perc_unassigned_transcripts

mean_tx <- mean(
    tbl$transcript_count,
    na.rm=TRUE
)

mean_gn <- mean(
    tbl$gene_count,
    na.rm=TRUE
)

cat(
    "\n",
    "#cells: ", num_cells,
    "; %unassigned: ", round(p_unassigned, 2), "\n",
    "mean #transcripts per cell: ", round(mean_tx, 2),
    "; mean #genes per cell: ", round(mean_gn, 2), "\n",
    "\n",
    sep=""
)
```

Approx. `r round(p_unassigned, 1)`% of transcripts were not assigned 
to any cell. The mean transcript count per cell is 
`r round(mean_tx, 1)`. The mean gene count per cell is `r round(mean_gn, 1)`. It is also relevant to inspect the mean transcript count per gene per cell. 

Let's write a function to plot features per cell type as boxplots:
```{r feature-by-celltype-box}
# use same color mapping as in python
pal <- py$col_celltype

feature_by_celltype_boxplot <- function(feature) {

    df <- as.data.frame(colData(se)) |>
        filter(
            !is.na(transferred_cell_type),
            !is.na(.data[[feature]])
        ) |>
        mutate(
            transferred_cell_type = reorder(
                transferred_cell_type,
                .data[[feature]],
                FUN = median,
                na.rm = TRUE
            )
        )

    ggplot(
        df,
        aes(
            x=transferred_cell_type,
            y=.data[[feature]],
            fill=transferred_cell_type
        )
    ) +
        geom_boxplot(outlier.size=0.4) +
        scale_fill_manual(
            values=pal,
            na.value="#808080"
        ) +
        labs(
            x="Cell type",
            y=feature
        ) +
        theme_bw() +
        theme(
            axis.text.x=element_text(angle=45, hjust=1),
            legend.position="none"
        )
}
```

On average, cells express only 1–2 transcripts per gene:

```{r plt-mean-transcripts-per-gene-box, fig.width=7, fig.height=4}
feature_by_celltype_boxplot("mean_transcripts_per_gene")
```

We can also inspect the transcript_density. Some workflows (see below) recommend filtering cells based on transcript density. 
However, this can bias filtering towards specific cell types, such as 
macrophages, mast cells, or dendritic cells, which often exhibit large 
cell areas but comparatively low transcript counts.

```{r plt-mean-transcript-density, fig.width=7, fig.height=4}
feature_by_celltype_boxplot("transcript_density")
```

Now, let's recreate the python `spatialdata_plot` from above in R (Helena please help)

```{r segtraq-plot-r, fig.width=14, fig.height=8}
#| code-fold: true

p1 <- plotSpatialData() +
    plotImage(sd, "he_image") +
    ggtitle("H&E")

p2 <- plotSpatialData() +
    plotImage(sd, "morphology_focus", ch="DAPI", c="purple") +
    ggtitle("DAPI")

p3 <- plotSpatialData() +
    plotShape(sd, "cell_boundaries", col=NA, fill="transferred_cell_type") +
    ggtitle("Cell type")  +
    scale_fill_manual(values=pal, na.value="#808080") +
    ggtitle("Cell type") 

p4 <- plotSpatialData() +
    plotShape(sd, "cell_boundaries", col=NA, fill="elongation") +
    ggtitle("Elongation")

p5 <- plotSpatialData() +
    plotShape(sd, "cell_boundaries", col=NA, fill="transcript_count") +
    ggtitle("Transcript count")

p6 <- plotSpatialData() +
    plotShape(sd, "cell_boundaries", col=NA,
        fill="negative_marker_contamination_counts") +
    ggtitle("Negative marker (contamination) count")

ps <- list(p1, p2, p3, p4, p5, p6)

wrap_plots(ps, nrow=2) &
    theme_void() &
    theme(
        plot.title=element_text(
            hjust=0.5,
            size=9
        ),
        plot.margin=margin(1, 1, 1, 1),
        legend.position="right"
    )
```

Stromal cells show higher elongation, whereas T and B cells tend to have less elongated shapes:

```{r plt-elongation-box, fig.width=7, fig.height=4}
feature_by_celltype_boxplot("elongation")
```

After inspecting baseline metrics, we next assess potential contamination.
Here, `SegTraQ` summarizes directed contamination between source and target
cell types based on negative marker expression.

NaN values indicate cell type pairs for which contamination cannot be assessed,
either because the cell types are not spatial neighbors or because they do not
share informative positive/negative marker combinations.

The dataset contains only
`r round(mean(tbl$transferred_cell_type == "tumor", na.rm=TRUE) * 100, 2)`
% tumor cells, which likely explains why little or no contamination originating
from tumor cells is detected.

About 19% of myoepithelial cells appear contaminated by DCIS2 cells,
which are common spatial neighbors. 3% of T cells are contaminated by
B cells. 

```{r plt-contamination-heatmap, fig.width=5, fig.height=4}

mat <- metadata(table(sd))$negative_marker_contamination_binary

df <- as.data.frame(mat) |>
    tibble::rownames_to_column("source_cell_type") |>
    pivot_longer(
        -source_cell_type,
        names_to="target_cell_type",
        values_to="contamination"
    )

ggplot(df, aes(
    x=target_cell_type,
    y=source_cell_type,
    fill=contamination
)) +
    geom_tile() +
    geom_text(aes(label=sprintf("%.2f", contamination)), size=2) +
    scale_fill_gradient(low="white", high="red") +
    labs(
        x="Target cell type",
        y="Source cell type",
        fill=NULL
    ) +
    theme_bw() +
    theme(
        axis.text.x=element_text(angle=45, hjust=1),
        panel.grid=element_blank()
    )
```

For demonstration, let's inspect mutually exclusive markers between B and T
cells. Here, we recover well-known B cell markers such as `MS4A1` and `CD19`,
which are not expected to be expressed in T cells.

```{r markers-intersection}
mk <- py$markers

B_pos <- unlist(mk$B$positive)
T_neg <- unlist(mk$T$negative)

mut_B_T <- intersect(B_pos, T_neg)
mut_B_T
```

Let's visualize the expression of `MS4A1` in T & B cells. To this end, we crop the dataset to a region with a high density of B and T cells. Visually, `MS4A1` transcripts appear closer to the cell membrane in T cells
than in B cells, which may be consistent with local transcript spillover.
```{r crop-T-B}
bb <- list(
    xmin=14e3, xmax=16e3,
    ymin=45e3, ymax=47e3)
sq <- crop(sd, bb)
```

```{r plt-ms4a1-tcells}
plotSpatialData() +
    plotShape(sq, "cell_boundaries", col=NA, fill="transferred_cell_type", alpha=0.8) +
    ggtitle("Cell type")  +
    scale_fill_manual(values=pal, na.value="#808080") +
    ggtitle("Cell type") +
    plotPoint(
        sq,
        key="MS4A1",
        colour="feature_name",
        col="black",
        size=0.3
    ) +
    ggtitle("MS4A1 transcripts in T and B cells")
```

Let's evaluate this with `SegTraQ`, which provides the point statistics (`ps`) module to quantify
subcellular transcript distributions within cells. For example, we can
compute transcript distances to the cell membrane either across all genes
and cells or for selected genes and cell types. For this, let's briefly jump back into Python.

```{python ps-module}
_ = st.ps.distance_to_membrane(
    genes=["MS4A1"], 
    cell_type_key="transferred_cell_type",
    cell_type_query=["B", "T"])
```

Let's jump back to R and plot the distance of `MS4A1` in B and T cells. 

```{r transfer-python-table-2}
se <- py$adata$as_SingleCellExperiment()

# remove old table
sd <- sd[c("images", "points", "shapes")]
(sd <- setTable(sd, y=se, 
    i="cell_boundaries", 
    ik="cell_id"))
```

Indeed, `MS4A1` is closer to the cell membrane in T than in B cells. This supports the interpretation that these signals likely arise from spillover rather than true biological expression.

```{r plt-ms4a1-dist, fig.width=4, fig.height=4}
df <- as.data.frame(colData(se)) |>
    dplyr::filter(
        transferred_cell_type %in% c("B", "T"),
        !is.na(distance_to_cell_membrane_norm_MS4A1)
    )

ggplot(
    df,
    aes(
        x=transferred_cell_type,
        y=distance_to_cell_membrane_norm_MS4A1,
        fill=transferred_cell_type
    )
) +
    geom_boxplot(outlier.size=0.4, alpha=0.7) +
    geom_jitter(
        aes(color=transferred_cell_type),
        width=0.15,
        size=0.8,
        alpha=0.5
    ) +
    scale_fill_manual(values=pal) +
    scale_color_manual(values=pal) +
    labs(
        x="Cell type",
        y="Distance to cell membrane"
    ) +
    theme_bw() +
    theme(legend.position="none")
```

To explore additional `SegTraQ` metrics, please refer to the
[SegTraQ documentation](https://lazdaria.github.io/SegTraQ/?utm_source=chatgpt.com).

`SegTraQ` can support the identification and filtering of low-quality cells
or samples based on morphological abnormalities, low transcript counts, or
increased contamination. It can also be used to benchmark segmentation methods
and help identify the most suitable segmentation strategy for a given dataset.

`SegTraQ` is still under active development, and additional metrics and
improvements will be introduced in future releases.

### other QC packages

The `SpatialData` object contains a single `table` element -- represented as a
`r BiocStyle::Biocpkg("SingleCellExperiment")` -- annotating `cell_boundaries`.
This `table` contains the single-cell level data: a gene $\times$ cell count 
matrix (`assay` slot `"X"`), gene/cell metadata (`row/colData`), as well as
spatial coordinates (`reducedDim` slot `"spatial"`).

```{r get-tbl}
(se <- sd$tables$cell_boundaries_table)
```

Note that these data originally stem from a Python `anndata` object written 
to `.zarr`, which were read into R using `r BiocStyle::Biocpkg("anndataR")`.
We can therefore perform standard processing tasks using tools available 
in R/Bioc (e.g., `r BiocStyle::Biocpkg("scrapper")`) such as filtering based on transcript_density:

```{r fil}
# exclude cells <2 MADs in log-counts per area
cpa <- log(se$transcript_counts/se$cell_area)
mad <- mad(cpa, na.rm=TRUE)
med <- median(cpa, na.rm=TRUE)
se$ex <- is.na(cpa) | (cpa < med-2*mad) 
```

```{r message=FALSE}
#| code-fold: true
plotSpatialData() + 
    plotShape(`table<-`(sd, value=se), col=NA, fill="ex") + 
    scale_fill_manual("exclude", values=c("grey", "red"))
```

We'll here filter out cells with comparatively low transcript density,
and proceed with processing steps standard for scRNA-seq data, namely:

- log-library size normalization
- highly variable gene (HVG) selection
- principal component analysis (PCA)
- shared-nearest neighbor (SNN) graph construction
- community detection using the Leiden algorithm

```{r pro}
base::table(se$ex)
se <- se[, !se$ex]
assayNames(se) <- "counts"
se <- normalizeRnaCounts.se(se)
se <- chooseRnaHvgs.se(se)
se <- runPca.se(se, features=rowData(se)$hvg)
se <- clusterGraph.se(se, method="leiden", resolution=0.5)
nk <- nlevels(se$clusters)
table(sd) <- se
```

Let's visualize the resulting cluster assignments spatially; from this, we can
observe that malignant regions are mostly made up of a single (large) cluster:

```{r plt-xy-clu, eval=FALSE, message=FALSE}
plotSpatialData() + 
    plotShape(sp, col=NA, fill="clusters") +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

### downstream

#### structures

```{r message=FALSE, warning=FALSE}
library(sosta)
library(spatstat.geom)
```

We'll start by construction a point pattern process (`ppp`) of 
cell centroid coordinates, using cluster assignments as `marks`:

```{r sosta-ppp, warning=FALSE}
xy <- centroids(SpatialData::transform(shape(sd)))
ow <- owin(xrange=range(xy[, 1]), yrange=range(xy[, 2]))
pp <- ppp(x=xy[, 1], y=xy[, 2], window=ow, marks=se$clusters)
```

We can now use cellular density (more precisely, the `ppp`'s intensity)
to identify multi-cellular anatomical substructures using `sosta`;
we'll then assign these to our object as a new `shape` element:

```{r sosta-roi, message=FALSE}
use <- which.max(table(se$clusters))
roi <- reconstructShapeDensity(pp, markSelect=use, dim=400)
st_geometry(roi) <- "geometry"
roi <- SpatialDataShape(roi)
shape(sd, "roi") <- roi
```

```{r sosta-plt, message=FALSE}
#| code-fold: true
plotSpatialData() + 
    plotShape(sd, col=NA, fill="clusters") + 
    plotShape(sd, "roi", col="red", fill=NA, linewidth=1) +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

Next, we can use regions to aggregate cell-level measurements into 
pseudo-bulks using `mask()`; here, `how` specifies how to do that:

```{r}
sd <- mask(sd, i="cell_boundaries", j="roi", how="sum")
(pb <- tail(tables(sd), 1)[[1]])
```

```{r}
pb <- normalizeRnaCounts.se(pb, 
    log=FALSE, 
    assay.type="sum",
    output.name="normcounts")
pc <- runPca(assay(pb, "normcounts"))
```

```{r}
# top-loaded features for PCs 1-3
apply(pc$rotation[, 1:3], 2, \(.) head(names(sort(abs(.), decreasing=TRUE)), 10))
```

```{r}
plotSpatialData() + 
    plotShape(sd, col=NA, fill="XBP1", assay="logcounts") +
    scale_fill_gradientn(colors=rev(hcl.colors(9, "Roma")))
```

AI tells us *"X-box binding protein 1 (XBP1) is a key transcription factor, 
activated by the IRE1/XBP1 branch of the Unfolded Protein Response (UPR), 
which promotes breast cancer development, progression, and resistance to 
therapy."*; so let's believe that and be proud.

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
