## CSAMA

In this demo, we'll be analyzing 5K-plex Xenium (10x Genomics) 
data on a tissue section from a human breast cancer biopsy, 
derived from [here](https://www.10xgenomics.com/datasets/xenium-prime-ffpe-human-breast-cancer).
These include the following elements:

- two `image`s (histology and immunofluorescence)
- two `shape`s (nuclei and membrane boundaries)
- one `table` (gene $\times$ cell data)
- one `point` (RNA transcripts)

A size-reduced example dataset has been deposited on Zenodo. 
We'll first retrieve the compressed *.zarr* store programatically, 
decompress it, and stash the data's on-disk file path for later:

```{r load-zarr}
# retrieve .zarr
url <- file.path(
    "https://zenodo.org/records/20120907/files",
    "BC_xenium_sdata_AL_LR.zarr.zip?download=1")
zip <- tempfile("sd", fileext=".zarr.zip")
download.file(url, zip, quiet=TRUE)

# decompress, stash & view contents
nm <- unzip(zip, exdir=dirname(zip))
zs <- grepv("zarr$", dirname(nm))
cat(basename(zs))
```

```{r show-zarr, echo=FALSE}
htmltools::pre(
    style="max-height:333px; overflow-y:auto;",
    paste(capture.output(fs::dir_tree(zs, recurse=2))[-1], collapse="\n"))
```

### dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(anndataR)
library(scrapper)
library(patchwork)
library(reticulate)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
library(SingleCellExperiment)
```

### setup

We can instruct our session to use an existing environment with our Python 
dependencies (here, `SegTraQ`) as follows. Note that `reticulate` provides 
similar commands to, e.g., use a virtual (not conda) environemnt instead:

::: {.column-margin}
The environment used here has been set up previously outside of R, 
from the command line, using these commands:
```
conda create -n segtraq python=3.11
conda activate segtraq
pip install segtraq
```
:::

```{r py-env}
# path to conda binary
bin <- "~/software/mambaforge/bin/conda" 
options(reticulate.conda_binary=bin)

# activate environment
use_condaenv("segtraq")
```

```{python py-opt, echo=FALSE}
import joblib
import logging
import warnings

warnings.filterwarnings("ignore")
joblib.parallel.DEFAULT_BACKEND = "threading"

logging.getLogger("spatialdata").propagate = False
logging.getLogger("spatialdata").setLevel(logging.ERROR)

logging.getLogger("spatialdata_plot").propagate = False
logging.getLogger("spatialdata_plot").setLevel(logging.ERROR)
```

### segmentation

```{python segtraq-load}
import spatialdata as sd
sdata = sd.read_zarr(r.zs)
sdata
```

```{python segtraq-calc}
import segtraq

sdata.shapes["cell_boundaries"].index.name = "cell_id"
sdata.shapes["nucleus_boundaries"].index.name = "cell_id"

st = segtraq.SegTraQ(sdata,
    tables_centroid_x_key=None,
    tables_centroid_y_key=None,
    images_key="morphology_focus")

st.run_baseline()

adata = st.sdata.tables["table"]
adata
```

```{python segtraq-plot}
#| code-fold: true
import spatialdata_plot
import matplotlib.pyplot as plt

fg, ax = plt.subplots(1, 3, figsize=(14, 4), constrained_layout=True)
fg.subplots_adjust(wspace=0.5)
ax = ax.flatten()

# plot histopathology
st.sdata.pl.render_images(
    "he_image"
).pl.show(
    ax=ax[0], 
    title="H&E")

# plot cell boundaries
st.sdata.pl.render_shapes(
    element="cell_boundaries",
    color="transcript_counts",
    method="matplotlib"
).pl.show(
    ax=ax[1], 
    colorbar=True,
    title="Tx counts")
    
# plot immunofluorescence
st.sdata.pl.render_images(
    "morphology_focus", 
    channel="DAPI"
).pl.show(
    ax=ax[2], 
    title="DAPI")
```

```{r echo=FALSE}
knitr::opts_chunk$set(eval=TRUE)
```

### exploratory

Let's load up the data in R, skipping the `table` layer. We then retrieve the 
`table`, including `SegTraQ` results, from Python, and set it to annotate cell
segmentation boundaries using `setTable()`; `ik` specifies the `spatialdata`'s
`instance_key` used to link observations between `table` and spatial element.

```{r}
# intentionally skip reading table layer; 
# we'll get it from Python with 'anndataR'
sd <- readSpatialData(zs, tables=FALSE)

# bug? shapes are missing this...
shape(sd) <- mutate(shape(sd), cell_id=`__index_level_0__`)
shape(sd)@meta$spatialdata_attrs$instance_key <- "cell_id"

# set Python table to annotate 
# cell segmentation boundaries
se <- py$adata$as_SingleCellExperiment()
(sd <- setTable(sd, y=se, 
    i="cell_boundaries", 
    ik="cell_id"))
```

#### histopathology

Let's start out by visualization the hematoxylin and eosin (H&E) staining.
H\&E provides a high-contrast visualization of tissue microarchitecture and 
cellular morphology through differential staining of nuclear and cytoplasmic 
components. To date, whole-slide H\&E images (WSIs) remain the clinical gold 
standard for cancer diagnosis, including tumor grading, staging, and prognosis.

```{r plt-he}
plotSpatialData() + plotImage(sd)
```

We can also zoom into a particular region by specifying a bounding box and
using `crop()` to subset the data accordingly:

```{r plt-he-box}
bb <- list(
    xmin=15e3, xmax=18e3, 
    ymin=44e3, ymax=46e3)
sp <- crop(sd, bb)
plotSpatialData() + plotImage(sp) 
```

#### immunofluorescence

In addition to the H\&E, the dataset includes 4 immunofluorescence (IF) markers:

* DAPI (nuclei) marks all cell nuclei and is used for cell segmentation
* ATP1A1/CD45/E-Cadherin (membrane) marks cell boundaries and major lineages
* 18S (cytoplasm) marks the cell body/cytoplasmic RNA and overall cell extent
* αSMA/Vimentin (stroma) marks mesenchymal/stromal tissue, including fibroblasts

```{r if-chs}
img <- image(sd, 2)
as.vector(channels(img))
```

The underlying image is thus a 3D array where the first dimension are channels, 
and the second and third dimensions are height and width; we can verify this
by viewing the `axes()` specified in the *.zattrs* associated with the image:

```{r if-dim}
do.call(rbind, axes(img))
cat(dim(img))
```

Let's visualize the four-plex composite image overlaying these markers;
for consistent visualization, we also specify colors to use for each channel:

```{r plt-if-all}
pal <- c("blue", "green", "cyan", "magenta")
chs <- names(pal) <- channels(img)
plotSpatialData() + plotImage(sd, 2, ch=chs, c=pal[chs])
```

Alternatively, we can visualize each channel separately; 
here, using the previously defined smaller region:

```{r plt-if-one}
ps <- lapply(chs, \(.) plotSpatialData() + plotImage(sp, 2, ch=., c=pal[.]) + ggtitle(.))
wrap_plots(ps, nrow=2) & theme_void() & theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

#### segmentation

We can also zoom in further to inspect how well cell segmentation boundaries 
align with histopathology and the immunofluorescence signals used to get them:

```{r plt-if-seg, fig.width=11, fig.height=6}
bb <- list(
    xmin=15e3, xmax=17e3,
    ymin=42e3, ymax=45e3)
sq <- crop(sd, bb)
plotSpatialData() + 
    plotImage(sq, 1) +
    plotShape(sq, fill=NA, col="black", linewidth=1/5) +
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, fill=NA, col="white", linewidth=1/5)
```

We can also go a step further, visualizing both cell nuclei (in gold) 
and membrane (in grey) boundaries togeher on an ever smaller region:

```{r}
bb <- list(
    xmin=15e3, xmax=16e3,
    ymin=43e3, ymax=44e3)
sq <- crop(sd, bb)
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, 1, fill=NA, col="grey", linewidth=1/3) +
    plotShape(sq, 2, fill=NA, col="gold", linewidth=1/3)
```

#### transcripts 

Going all the way, let's zoom in to a couple cells in order to visualize 
4,000 individual transcript molecules as well; we'll color these by whether
or not they have been determined to `overlaps_nucleus` (according to 10x):

```{r}
bb <- list(
    xmin=15200, xmax=15800,
    ymin=43200, ymax=43800)
sq <- crop(sd, bb)
point(sq) <- mutate(point(sq), overlaps_nucleus=as.logical(overlaps_nucleus))
plotSpatialData() + 
    plotImage(sq, 2, ch=chs, c=pal[chs]) +
    plotShape(sq, 1, fill=NA, col="grey", linewidth=1/3) +
    plotShape(sq, 2, fill=NA, col="gold", linewidth=1/3) +
    plotPoint(sq, n=4e3, col="overlaps_nucleus", size=0) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    scale_color_manual(values=c("grey", "red"))
```

Going back to a larger scale...

```{r fig.width=8, fig.height=4}
#| code-fold: true
bb <- list(xmin=16e3, xmax=18e3, ymin=45e3, ymax=46e3)
sq <- crop(sd[c("images", "points"), list(2, 1)], bb)
plotSpatialData() + 
    plotImage(sq, ch="DAPI", c="purple") +
    plotPoint(sq, n=1e4, col="ivory", size=0) +
    theme_void() + theme(legend.position="none")
```

Let's quantify the number of transcripts that do no fall within any cell. 
We do this by `mask`ing `transcripts` by the `cell_boundaries` shape element; 
this will return a new `SpatialData` object with an additional table in which
rows = features and columns = cells; here, one extra column ("0") will collect 
data on transcripts that do not intersect with *any* shape:

```{r}
se <- mask(sd, 
    i="transcripts", j="cell_boundaries")$
    tables$transcripts_by_cell_boundaries
# unassigned
(n <- se$n_instances[["0"]]) 
# percentage
m <- length(point(sd))
round(100*n/m, 2)
```

### processing

The `SpatialData` object contains a single `table` element -- represented as a
`r BiocStyle::Biocpkg("SingleCellExperiment")` -- annotating `cell_boundaries`.
This `table` contains the single-cell level data: a gene $\times$ cell count 
matrix (`assay` slot `"X"`), gene/cell metadata (`row/colData`), as well as
spatial coordinates (`reducedDim` slot `"spatial"`).

```{r get-tbl}
(se <- sd$tables$cell_boundaries_table)
```

Note that these data originally stem from a Python `anndata` object written 
to `.zarr`, which were read into R using `r BiocStyle::Biocpkg("anndataR")`.
We can therefore perform standard processing tasks using tools available 
in R/Bioc (e.g., `r BiocStyle::Biocpkg("scrapper")`) such as quality control:

```{r fil}
# exclude cells <2 MADs in log-counts per area
cpa <- log(se$transcript_counts/se$cell_area)
mad <- mad(cpa, na.rm=TRUE)
med <- median(cpa, na.rm=TRUE)
se$ex <- is.na(cpa) | (cpa < med-2*mad) 
```

```{r message=FALSE}
#| code-fold: true
plotSpatialData() + 
    plotShape(`table<-`(sd, value=se), col=NA, fill="ex") + 
    scale_fill_manual("exclude", values=c("grey", "red"))
```

We'll here filter out cells with comparatively low transcript density,
and proceed with processing steps standard for scRNA-seq data, namely:

- log-library size normalization
- highly variable gene (HVG) selection
- principal component analysis (PCA)
- shared-nearest neighbor (SNN) graph construction
- community detection using the Leiden algorithm

```{r pro}
base::table(se$ex)
se <- se[, !se$ex]
assayNames(se) <- "counts"
se <- normalizeRnaCounts.se(se)
se <- chooseRnaHvgs.se(se)
se <- runPca.se(se, features=rowData(se)$hvg)
se <- clusterGraph.se(se, method="leiden", resolution=0.5)
nk <- nlevels(se$clusters)
table(sd) <- se
```

Let's visualize the resulting cluster assignments spatially; from this, we can
observe that malignant regions are mostly made up of a single (large) cluster:

```{r plt-xy-clu, eval=FALSE, message=FALSE}
plotSpatialData() + 
    plotShape(sp, col=NA, fill="clusters") +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

### downstream

#### structures

```{r message=FALSE, warning=FALSE}
library(sosta)
library(spatstat.geom)
```

We'll start by construction a point pattern process (`ppp`) of 
cell centroid coordinates, using cluster assignments as `marks`:

```{r sosta-ppp, warning=FALSE}
xy <- centroids(SpatialData::transform(shape(sd)))
ow <- owin(xrange=range(xy[, 1]), yrange=range(xy[, 2]))
pp <- ppp(x=xy[, 1], y=xy[, 2], window=ow, marks=se$clusters)
```

We can now use cellular density (more precisely, the `ppp`'s intensity)
to identify multi-cellular anatomical substructures using `sosta`;
we'll then assign these to our object as a new `shape` element:

```{r sosta-roi, message=FALSE}
use <- which.max(table(se$clusters))
roi <- reconstructShapeDensity(pp, markSelect=use, dim=400)
st_geometry(roi) <- "geometry"
roi <- SpatialDataShape(roi)
shape(sd, "roi") <- roi
```

```{r sosta-plt, message=FALSE}
#| code-fold: true
plotSpatialData() + 
    plotShape(sd, col=NA, fill="clusters") + 
    plotShape(sd, "roi", col="red", fill=NA, linewidth=1) +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

Next, we can use regions to aggregate cell-level measurements into 
pseudo-bulks using `mask()`; here, `how` specifies how to do that:

```{r}
sd <- mask(sd, i="cell_boundaries", j="roi", how="sum")
(pb <- tail(tables(sd), 1)[[1]])
```

```{r}
pb <- normalizeRnaCounts.se(pb, 
    log=FALSE, 
    assay.type="sum",
    output.name="normcounts")
pc <- runPca(assay(pb, "normcounts"))
```

```{r}
# top-loaded features for PCs 1-3
apply(pc$rotation[, 1:3], 2, \(.) head(names(sort(abs(.), decreasing=TRUE)), 10))
```

```{r}
plotSpatialData() + 
    plotShape(sd, col=NA, fill="XBP1", assay="logcounts") +
    scale_fill_gradientn(colors=rev(hcl.colors(9, "Roma")))
```

AI tells us *"X-box binding protein 1 (XBP1) is a key transcription factor, 
activated by the IRE1/XBP1 branch of the Unfolded Protein Response (UPR), 
which promotes breast cancer development, progression, and resistance to 
therapy."*; so let's believe that and be proud.

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
